---
title: "Machine Learning Exercise"
author: "Bill Eddins"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: true
---
```{r SetUp, echo=FALSE, message=FALSE,warning=FALSE}
require(ggplot2); require(caret); rm(list=ls()); set.seed(1952)
knitr::opts_chunk$set(fig.path='figures/')
```
#Overview

The purpose of this report is to get credit for an online course in machine learning. See the following link for more information. Look for the course titled 'Practical Machine Learning'.

[Data Science Specialization](https://www.coursera.org/specialization/jhudatascience/1?utm_source=catalog)

The data come from a reseach study of subjects conducting physical activites using a barbell while being monitored by wearable devices. See the following link.  

[Human Activity Recognition](http://groupware.les.inf.puc-rio.br/har)


##A machine learning algorithm predicting activity quality

All R code for this report can be found in the appendix. The outcome of the study is to build a model that predicts barbell activies using a testing file and a training file. The response variable is 'classe' which designates a row as being one of the five classes of barbell activities. In order to reduce out of sample errors, the testing file is partitioned into training, validating and testing data sets. Then, the training file which has 20 records will be used to predict the type of exercise.  

Training consists of building three classification models using a generalized boosted model, a random forest model, and a treebag model. The varImp was generated on each model to determine the variables which have the most impact on the response variable. Validating consists of predicting classe membership for each of the classification models and then picking the best model by use of a confusionMatrix. The reader should note that the random forest model is the best one to use for testing since it has Accuracy equal to 0.9829.  

Testing will narrow down the list of predictors to the 10 predictors that were identified by the impVar function during validating. Those predictors include: roll_ belt, yaw_belt, pitch_belt, magnet_dumbbell_y, magnet_dumbbell_z, pitch_forearm, roll_forearm, accel_dumbbell_y, roll_dumbbell, and magnet_belt_y. 

##Conclusions

This report has controlled for out of sample error by two techniques. First, train control parameters were passed to the train method to do cross validation. Second, a three phased approach to constructing the final model was employed including a training, validating, and testing phases. Then, 10 variables were identified and were fitted during the testing phase. After that, the testing file with twenty records was input and the type of barbell activity was predicted.

If I get time, I will come back, discuss more and fit some kind of graphs. 

##Appendix  
```{r LoadTraining, echo=TRUE, message=FALSE,warning=FALSE}
chrFile = "pml-training.csv"; curls <- read.csv(chrFile, header=TRUE)
dim(curls)
```
```{r CleanUpTraining, echo=TRUE, message=FALSE,warning=FALSE}
#Remove unnecessary columns (All NAs, calced cols, and unrelated cols)
curls <- curls[, -c(1:7, 12:36, 50:59, 69:83, 87:101, 103:112, 125:139, 141:150)]
curls[is.na(curls)] <- 0
prop.table(table(curls$classe))
dim(curls)
```
```{r SplitTrainingDataSet, echo=TRUE, message=FALSE,warning=FALSE}
#See following url for more info on this technique
#http://amunategui.github.io/blending-models/
curls <- curls[sample(nrow(curls)),]; split <- floor(nrow(curls)/3)
training <- curls[0:split,]; validating <- curls[(split+1):(split*2),]
testing <- curls[(split*2+1):nrow(curls),]; labelName <- 'classe'
```
```{r PerformModeling, echo=TRUE, message=FALSE,warning=FALSE}
#Create list of predictors, cv control and three models (gbm, rf and treebag)
predictors <- names(training)[names(training) != labelName]
myControl <- trainControl(method = "cv", number = 3, repeats = 1, returnResamp = "none")
mod.gb <- train(training[,predictors], training[,labelName], method='gbm', 
        trControl=myControl, verbose=FALSE)
mod.rf <- train(training[,predictors], training[,labelName], method='rf', trControl=myControl)
mod.tb <- train(training[,predictors], training[,labelName], method='treebag',
        trControl=myControl)
gb.vars <- varImp(mod.gb); rf.vars <- varImp(mod.rf); tb.vars <- varImp(mod.tb)
the.vars <- rbind(gb.vars$importance, rf.vars$importance, tb.vars$importance)
write.csv(the.vars,"the.vars.csv")
```                     
```{r PerformValidating, echo=TRUE, message=FALSE,warning=FALSE}
#Predict classe membership using the validating data
validating$prob.gb <- predict(object=mod.gb, validating[,predictors])
validating$prob.rf <- predict(object=mod.rf, validating[,predictors])
validating$prob.tb <- predict(object=mod.tb, validating[,predictors])
#Do matrix against original data
cm.gb <- confusionMatrix(validating$prob.gb, validating$classe) #Accuracy : 0.9561
cm.rf <- confusionMatrix(validating$prob.rf, validating$classe) #Accuracy : 0.9829 Winner
cm.tb <- confusionMatrix(validating$prob.tb, validating$classe) #Accuracy : 0.9657
print("gbm"); cm.gb$overall; print("rf"); cm.rf$overall; print("treebag"); cm.tb$overall
```
```{r PerformTesting, echo=TRUE, message=FALSE,warning=FALSE}
mod.fin <- train(classe ~ roll_belt + yaw_belt + pitch_belt + magnet_dumbbell_y
        + magnet_dumbbell_z + pitch_forearm + roll_forearm + magnet_dumbbell_y 
        + roll_dumbbell + magnet_belt_y, method='rf', trControl=myControl, data = testing)
testing$prob.rf <- predict(object=mod.fin, testing)
confusionMatrix(testing$classe, testing$prob.rf) #Accuracy = 1
mod.fin$results
```
```{r MakeFinalPredictions, echo=TRUE, message=FALSE,warning=FALSE}
chrFile = "pml-testing.csv"; curls.testing <- read.csv(chrFile, header=TRUE)
curls.testing$pred <- predict(object=mod.fin, newdata = curls.testing)
answers <- curls.testing$pred
pml_write_files = function(x){n = length(x)
  for(i in 1:n){filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(answers)
```




